---
title: "Tidyverse Part 1"
author: "C. Rosemond"
date: "November 2, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Library

```{r library, message=FALSE}
library(tidyverse)
```

## Data Set(s)
I selected two fivethirtyeight data sets: one that contains current Soccer Power Index (SPI) ratings and rankings for men's club teams and a second that contains match-by-match SPI ratings and forecasts back to 2016.

URL: https://github.com/fivethirtyeight/data/tree/master/soccer-spi

## readr - read_csv()
The readr package facilitates the reading in of 'rectangular' data like .csv files or other delimited files. Here, I use the read_csv() function to read in two data sets: the global rankings, or 'rankings', and the matches, or 'matches'.

```{r read_csv1, eval=TRUE, message=FALSE}
rankings <- read_csv('https://projects.fivethirtyeight.com/soccer-api/club/spi_global_rankings.csv')
matches <- read_csv('https://projects.fivethirtyeight.com/soccer-api/club/spi_matches.csv')
head(rankings)
tail(matches)
```


## dplyr - mutate()
The dplyr package provides a grammar for the manipulation of data--notably, in data frames or tibbles. Here, I use the mutate function to add a new column--a match ID--to the matches tibble.

```{r mutate, eval=TRUE}
matches <- mutate(matches, match = row_number())
matches <- matches[,c(23,1:22)]
head(matches)
```


## dplyr - select()
The select function from dplyr enables the selection of data frame columns by name or helper function. Here, I select and keep the first six columns ('match' through 'team2') from the matches tibble.

```{r select, eval=TRUE}
matches <- select(matches, match:team2)
head(matches)
```


## dplyr - filter()
The filter function from dplyr enables the subsetting of rows based on specified logical criteria. Here, I select matches that occurred from November 1st through November 7th.

```{r filter, eval=TRUE}
matches <- filter(matches, date >= '2019-11-01' & date <= '2019-11-07')
head(matches)
```


## tidyr - gather()
The tidyr package is designed to facilitate reshaping data. Here, I use the gather() function to reshape the matches tibble from wide to long format, gathering the separate team columns.

```{r gather, eval=TRUE}
matches <- matches %>% gather(-match, -date, -league_id, -league, key=team_number, value=name) %>% select(-team_number)
head(matches)
```


## dplyr - arrange()
The arrange function from dplyr enables the sorting of data based upon column values. Here, I arrange the matches tibble by match number.

```{r arrange, eval=TRUE}
matches <- arrange(matches, match)
head(matches)
```


## dplyr - left_join()
The left_join function works similarly to its SQL counterparts. I finish by using 'name' to merge the matches tibble with the rankings tibble, which contains club rankings and ratings as of November 7th.

```{r left_join, eval=TRUE}
merged <- dplyr::left_join(matches, rankings, by='name')
merged <- select(merged, -league.y, - off, -def)
head(merged)
```


## Extended by Mario Pena


## dplyr::count()

### Usage: count(x, ..., wt = NULL, sort = FALSE)

**Count number of rows in each group defined by the variables in... Also tally()**. *count(Iris, Species)*
    
**Example**

This function will help count the number of matches in each league, however, we add the mutate function to divide the number by 2 because the count includes the two teams in each match:

```{r}
merged_count <- merged %>% count(league.x) %>% mutate(matches = n/2)
merged_count
```

## ggplot2::ggplot()

### Usage: ggplot(data = DATA) + GEOM_FUNCTION(mapping = aes (MAPPING), stat = sTAT, position = POSITION) + COORDINATE_FUNCTION + FACET_FUNCTION + SCALE_FUNCTION + THEME_FUNCTION

**Use data frame as input for a graphic and specify the set of plot aesthetics intended to be common throughout all subsequent layers unless specifically overridden.** *ggplot(data = mpg, aes(x = city, y = hwy)) Begins a plot that you finis by adding layers to.*
    
**Example**

The ggplot() function below is used to visualize the count of matches in each league. I have added different layers such as title, and labels:

```{r}
ggplot(merged_count, aes(x=reorder(league.x, -matches), matches)) + geom_bar(stat="identity", width = 0.5, fill = "tomato2") + labs(x = "Leagues", y = "Count", title = "League Matches") + theme(axis.text.x = element_text(angle = 60, hjust = 1, size=8))
```
